---
title: "Converting hand-smoothed norms tables into print-format output"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Overview

The purpose of this script is to convert lookup tables from cNORM, that have been modified by a project director (e.g., hand-smoothing has been applied) into print-format lookup tables. Because there is a step of human-intervention outside cNORM, the process cannot be automated as part of the basic cNORM workflow.

The input tables embody a hierarchy of data in which there is one table per test, and each of these tables holds the lookup columns for all age strats. In addition, you lookup raw scores in the left-ward column and read to the right to find the associated standard scores. here is the head of one such table.

#### Executable Code

```{r print-format-norms, eval = FALSE}
suppressMessages(library(here))
suppressMessages(library(tidyverse))
suppressMessages(library(writexl))

input_test_names <- c("lske", "lswe", "rhme", "rlne", "sege", "snwe")
output_test_names <- c("LSK-E", "LSW-E", "RHY-E", "RLN-E", "SEG-E", "SPW-E")
tod_form <- "TOD-E"
norm_type <- "age"
input_file_path <- "INPUT-FILES/PRINT-FORMAT-NORMS-TABLES/"
output_file_path <- "OUTPUT-FILES/PRINT-FORMAT-NORMS-TABLES/"

input_files_ta <- map(
  input_test_names,
  ~
  suppressMessages(read_csv(here(str_c(
  input_file_path, .x, "-", norm_type, ".csv"
))))
) %>% 
  set_names(input_test_names)

perc_ss_cols <- suppressMessages(read_csv(here(str_c(
  input_file_path, "perc-ss-cols.csv"
))))

age_strat <- input_files_ta[[1]] %>% 
  select(-raw) %>% 
  names()

print_lookups_ta <- input_files_ta %>%
  map(~
        .x %>% 
  pivot_longer(contains("-"), names_to = "age_strat", values_to = "ss") %>%
  arrange(age_strat) %>%
  group_by(age_strat) %>%
  complete(ss = 40:130) %>%
  group_by(age_strat, ss) %>%
  filter(n() == 1 | n() > 1 & row_number()  %in% c(1, n())) %>%
  summarize(raw = str_c(raw, collapse = '--')) %>%
  mutate(across(raw, ~ case_when(is.na(.x) ~ '-', TRUE ~ .x))) %>%
  arrange(age_strat, desc(ss)) %>%
  pivot_wider(names_from = age_strat,
              values_from = raw) %>%
  filter(!is.na(ss)) %>%
  right_join(perc_ss_cols, by = "ss") %>%
  relocate(perc, .before = "ss")
) %>% 
  set_names(input_test_names)

age_strat_cols_ta <-  print_lookups_ta %>%
  map( ~
         map(age_strat,
             ~
               .y %>%
               select(perc, ss,!!sym(.x)), .y = .x) %>%
         set_names(age_strat))

age_test_names_flat <- cross2(age_strat, input_test_names) %>% 
  map_chr(str_c, collapse = "_")

age_test_cols_flat <- flatten(age_strat_cols_ta) %>% 
  set_names(age_test_names_flat)

age_test_cols_at <- map(
  age_strat,
  ~
  keep(age_test_cols_flat, str_detect(names(age_test_cols_flat), .x))
)

print_lookups_at <- age_test_cols_at %>% 
  map(
    ~
      .x %>% 
      reduce(left_join, by = c("perc", "ss")) %>% 
      rename_with(~ output_test_names, contains("-"))
  ) %>% 
  set_names(age_strat)

write_xlsx(print_lookups_at,
           here(
             str_c(
               output_file_path, tod_form, "-print-lookup-tables-", norm_type, ".xlsx"
             ))
           )
```

#### Commented Snippets

Text.

```{r print-format-norms, echo = 1:6, eval = F}
```
